### 2024-03-26 docs/projects/geography/chicago-bikes/index.qmd
Chicago bikes - rerendered

Here, we used the [Noj](https://scicloj.github.io/noj/) wrapper
of [Hanami](https://github.com/jsa-aerial/hanami),
which makes it work nicely with Tablecloth.

@@

::: {.sourceClojure}
```clojure
(delay
  (-> processed-trips
      (tc/group-by [:hour])
      (tc/aggregate {:n tc/row-count})
      (tc/order-by [:hour])))
```
:::

::: {.clay-dataset}
_unnamed [24 2]:

| :hour |    :n |
|------:|------:|
|     0 |  5093 |
|     1 |  3227 |
|     2 |  1871 |
|     3 |  1162 |
|     4 |   948 |
|     5 |  2801 |
|     6 |  8859 |
|     7 | 17058 |
|     8 | 22162 |
|     9 | 17299 |
|   ... |   ... |
|    13 | 24807 |
|    14 | 25866 |
|    15 | 30699 |
|    16 | 40338 |
|    17 | 46854 |
|    18 | 38245 |
|    19 | 27078 |
|    20 | 17231 |
|    21 | 14367 |
|    22 | 10810 |
|    23 |  7317 |

@@

::: {.sourceClojure}
```clojure
(defn as-geo [vega-lite-spec]
  (-> vega-lite-spec
      (update :encoding update-keys (fn [k]
                                      (get {:x :latitude
                                            :y :longitude
                                            :x2 :latitude2
                                            :y2 :longitude2}
                                           k k)))
      (assoc :projection {:type :mercator})))
```
:::

::: {.sourceClojure}
```clojure
(delay
  (-> processed-trips
      (tc/random 1000 {:seed 1})
      (hanami/plot ht/rule-chart
                   {:X :start_lat
                    :Y :start_lng
                    :X2 :end_lat
                    :Y2 :end_lng})
      as-geo))
```
:::

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"encoding":{"longitude":{"field":"start_lng","type":"quantitative"},"latitude":{"field":"start_lat","type":"quantitative"},"longitude2":{"field":"end_lng","type":"quantitative"},"latitude2":{"field":"end_lat","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":"rule","width":400,"background":"floralwhite","projection":{"type":"mercator"},"height":300,"data":{"url":"index_files\/2.csv","format":{"type":"csv"}}});</script></div>
```

::: {.sourceClojure}
```clojure
(delay
  (-> processed-trips
      (tc/random 1000 {:seed 1})
      (hanami/plot ht/rule-chart
                   {:X :start-local-x
                    :Y :start-local-y
                    :X2 :end-local-x
                    :Y2 :end-local-y
                    :XSCALE {:zero false}
                    :YSCALE {:zero false}})))
```
:::

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"encoding":{"y":{"scale":{"zero":false},"field":"start-local-y","type":"quantitative"},"x":{"scale":{"zero":false},"field":"start-local-x","type":"quantitative"},"y2":{"scale":{"zero":false},"field":"end-local-y","type":"quantitative"},"x2":{"scale":{"zero":false},"field":"end-local-x","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":"rule","width":400,"background":"floralwhite","height":300,"data":{"url":"index_files\/2.csv","format":{"type":"csv"}}});</script></div>
```

::: {.sourceClojure}
```clojure
(delay
  (-> processed-trips
      (tc/random 1000 {:seed 1})
      (hanami/plot ht/layer-chart
                   {:TITLE "Chicago bike trips"
                    :LAYER [{:data {:url "notebooks/data/chicago.geojson"
                                    :format {:type "topojson"}}
                             :mark {:type "geoshape"
                                    :filled false
                                    :clip true
                                    :opacity 0.3}}
                            (as-geo
                             (hanami/plot nil
                                          ht/rule-chart
                                          {:X :start_lat
                                           :Y :start_lng
                                           :X2 :end_lat
                                           :Y2 :end_lng
                                           :OPACITY 0.1}))]})))
```
:::

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"usermeta":{"embedOptions":{"renderer":"svg"}},"width":400,"background":"floralwhite","title":{"text":"Chicago bike trips"},"layer":[{"data":{"url":"notebooks\/data\/chicago.geojson","format":{"type":"topojson"}},"mark":{"type":"geoshape","filled":false,"clip":true,"opacity":0.3}},{"encoding":{"longitude":{"field":"start_lng","type":"quantitative"},"opacity":{"value":0.1},"latitude":{"field":"start_lat","type":"quantitative"},"longitude2":{"field":"end_lng","type":"quantitative"},"latitude2":{"field":"end_lat","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":"rule","width":400,"background":"floralwhite","projection":{"type":"mercator"},"height":300}],"height":300,"data":{"url":"index_files\/2.csv","format":{"type":"csv"}}});</script></div>
```

@@

@@

Let us plot a few clusters:

::: {.sourceClojure}
```clojure
(defn as-png [vega-lite-spec]
  ;; to improve performance, we avoid rendering as SVG in this case:
  (-> vega-lite-spec
      (assoc :usermeta {:embedOptions {:renderer :png}})))
```
:::

::: {.sourceClojure}
```clojure
(delay
  (-> processed-trips
      (tc/add-column :cluster (:clustering clustering))
      (tc/group-by [:cluster])
      (tc/without-grouping->
       (tc/order-by (fn [ds]
                      (-> ds :data tc/row-count))
                    :desc)
       (tc/head 20))
      (tc/aggregate {:n tc/row-count
                     :map (fn [trips]
                            [(-> trips
                                 (hanami/plot ht/layer-chart
                                              {:TITLE "Chicago bike trips"
                                               :LAYER [{:data {:url "notebooks/data/chicago.geojson"
                                                               :format {:type "topojson"}}
                                                        :mark {:type "geoshape"
                                                               :filled false
                                                               :clip true
                                                               :opacity 0.3}}
                                                       (as-geo
                                                        (hanami/plot nil
                                                                     ht/rule-chart
                                                                     {:X :start_lat
                                                                      :Y :start_lng
                                                                      :X2 :end_lat
                                                                      :Y2 :end_lng
                                                                      :OPACITY 0.01}))
                                                       (as-geo
                                                        (hanami/plot nil
                                                                     ht/point-chart
                                                                     {:X :start_lat
                                                                      :Y :start_lng
                                                                      :MCOLOR "purple"
                                                                      :OPACITY 0.01}))
                                                       (as-geo
                                                        (hanami/plot nil
                                                                     ht/point-chart
                                                                     {:X :end_lat
                                                                      :Y :end_lng
                                                                      :MCOLOR "green"
                                                                      :OPACITY 0.01}))]})
                                 as-png)])
                     :hours (fn [trips]
                              [(hour-counts-plot trips)])})
      kind/table))
```
:::

>>>>

Here, we used the [Noj](https://scicloj.github.io/noj/) wrapper
of [Hanami](https://github.com/jsa-aerial/hanami),
which makes it compose conveniently with Tablecloth,
eventually generating [Vega-Lite](https://vega.github.io/) plots.

@@

### Time
Let us see how trip start hours are distributed along the day.

@@

### Space

We wish to plot our trips as line segments
(Vega-Lite [Rule](https://vega.github.io/vega-lite/docs/rule.html)s)
over a map.

Hanami's templates create Vega-Lite plots with `x` and `y` coordinates,
but we need geographical plots, with `latitude` and `longitude`.

The following function will allow us to convert typical Hanami plots
to geographical ones.

::: {.sourceClojure}
```clojure
(defn as-geo [vega-lite-spec]
  (-> vega-lite-spec
      (update :encoding update-keys (fn [k]
                                      (get {:x :latitude
                                            :y :longitude
                                            :x2 :latitude2
                                            :y2 :longitude2}
                                           k k)))
      (assoc :projection {:type :mercator}
             ;; rendering as PNG this time
             ;; to make this page more lightweight
             :usermeta {:embedOptions {:renderer :png}})))
```
:::

Let us try it out with a sample of points.

::: {.sourceClojure}
```clojure
(delay
  (-> processed-trips
      (tc/random 10000 {:seed 1})
      (hanami/plot ht/rule-chart
                   {:X :start_lat
                    :Y :start_lng
                    :X2 :end_lat
                    :Y2 :end_lng
                    :OPACITY 0.1})
      as-geo))
```
:::

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"encoding":{"longitude":{"field":"start_lng","type":"quantitative"},"opacity":{"value":0.1},"latitude":{"field":"start_lat","type":"quantitative"},"longitude2":{"field":"end_lng","type":"quantitative"},"latitude2":{"field":"end_lat","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"png"}},"mark":"rule","width":400,"background":"floralwhite","projection":{"type":"mercator"},"height":300,"data":{"url":"index_files\/2.csv","format":{"type":"csv"}}});</script></div>
```

Let us add the neighbourhood geometries as an additional layer.

::: {.sourceClojure}
```clojure
(delay
  (-> processed-trips
      (tc/random 10000 {:seed 1})
      (hanami/plot ht/layer-chart
                   {:TITLE "Chicago bike trips"
                    :LAYER [{:data {:url "notebooks/data/chicago.geojson"
                                    :format {:type "topojson"}}
                             :mark {:type "geoshape"
                                    :filled false
                                    :clip true
                                    :opacity 0.5}}
                            (as-geo
                             (hanami/plot nil ; use the global dataset for this layer
                                          ht/rule-chart
                                          {:X :start_lat
                                           :Y :start_lng
                                           :X2 :end_lat
                                           :Y2 :end_lng
                                           :OPACITY 0.1}))]})))
```
:::

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"usermeta":{"embedOptions":{"renderer":"svg"}},"width":400,"background":"floralwhite","title":{"text":"Chicago bike trips"},"layer":[{"data":{"url":"notebooks\/data\/chicago.geojson","format":{"type":"topojson"}},"mark":{"type":"geoshape","filled":false,"clip":true,"opacity":0.5}},{"encoding":{"longitude":{"field":"start_lng","type":"quantitative"},"opacity":{"value":0.1},"latitude":{"field":"start_lat","type":"quantitative"},"longitude2":{"field":"end_lng","type":"quantitative"},"latitude2":{"field":"end_lat","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"png"}},"mark":"rule","width":400,"background":"floralwhite","projection":{"type":"mercator"},"height":300}],"height":300,"data":{"url":"index_files\/2.csv","format":{"type":"csv"}}});</script></div>
```

@@

To recognize repeating patterns of similar trips,
we use [k-means clustering](https://en.wikipedia.org/wiki/K-means_clustering)
over trips, representing them as quadruple of points
using the plane coordinates of their start and end.

@@

Let us plot the biggest of the resulting clusters.
We will plot each cluster separately in time and in space as before.
To the line segments over the map we add the start and end points,
in purple and green, respectively.

::: {.sourceClojure}
```clojure
(delay
  (-> processed-trips
      (tc/add-column :cluster (:clustering clustering))
      (tc/group-by [:cluster])
      ;; sort the groups in ascending order by their size,
      ;; and take the first 20 (which are the largest ones)
      ;; (we use `without-grouping->` to order and take
      ;; across groups rather tan within groups):
      (tc/without-grouping->
       (tc/order-by (fn [ds]
                      (-> ds :data tc/row-count))
                    :desc)
       (tc/head 20))
      (tc/aggregate {:n tc/row-count
                     :map (fn [trips]
                            [(-> trips
                                 (hanami/plot ht/layer-chart
                                              {:TITLE "Chicago bike trips"
                                               :LAYER [{:data {:url "notebooks/data/chicago.geojson"
                                                               :format {:type "topojson"}}
                                                        :mark {:type "geoshape"
                                                               :filled false
                                                               :clip true
                                                               :opacity 0.3}}
                                                       (as-geo
                                                        (hanami/plot nil
                                                                     ht/rule-chart
                                                                     {:X :start_lat
                                                                      :Y :start_lng
                                                                      :X2 :end_lat
                                                                      :Y2 :end_lng
                                                                      :OPACITY 0.01}))
                                                       (as-geo
                                                        (hanami/plot nil
                                                                     ht/point-chart
                                                                     {:X :start_lat
                                                                      :Y :start_lng
                                                                      :MCOLOR "purple"
                                                                      :OPACITY 0.01}))
                                                       (as-geo
                                                        (hanami/plot nil
                                                                     ht/point-chart
                                                                     {:X :end_lat
                                                                      :Y :end_lng
                                                                      :MCOLOR "green"
                                                                      :OPACITY 0.01}))]}))])
                     :hours (fn [trips]
                              [(hour-counts-plot trips)])})
      kind/table))
```
:::
